---
title: 'Template - Lab 3: Exploring Happiness'
subtitle: "Due: Thursday 10/03/2019 by 10 pm"
author: "Lynna Chu"

date: "09/26/2019"
output: html_document
---


## Background

The General Social Survey (GSS) has been run by NORC every other year since 1972 to keep track of current opinions across the United States. We are working with an excerpt of the dataset that covers years 1972 to 2016. The variable names along with a brief description are as follows: 

| Variable      | Description   |    
| ------------- |:-------------:| 
| Year          | Year for this respondent | 
| Age           | Age of respondent | 
| Sex           | Sex of respondent     | 
|....           | .....      | 

Now it's important to understand what each column stands for. You can find a codebook with explanations for each of the variables at https://gssdataexplorer.norc.org/

## Data Cleaning 

Initial exploration revealed that some data processing was needed before we could proceed with further visualization. All missing values were encoded with 'NA' as follows:
```{r}
library(tidyverse) #loading tidyverse to use dplyr functions
#importing the code from Lab3.Rmd to initialize happy, and make names lowercase
happy <- readRDS("HAPPY.rds") #initializing the Happy dataframe
names(happy) <- tolower(names(happy)) #making all column headings lowercase for ease of usability 

head(happy) #checking first 6 records


happy
(happy[is.na(happy) | happy == "IAP" | happy == "DK"] = NA) # Here we identified the na values and the values "IAP" and "DK" and inserted "NA"
happy

table(is.na(happy)) # used table() to view how many NA values there were in the Dataset, we saw there to be 36,391 instances of NA across ALL COLUMNS.

ggplot(happy, aes(x=happy)) + geom_bar() + labs(title = "Bar Plot of happiness Index Distribution", x = "Happiness Type", y = "Frequency") #this barchart shows the distribution of the happiness index, as well as NA values in the happy column.

```

In total there were `36391 ` number of missing values for the following `11` variables. 

Variables were also cast into the correct type in order to ease data visualization. This was done by:
```{r}
str(happy) # Looked at the variables' data types to find that age should be converted to Integer

happy = mutate(happy, age = as.integer(age)) # Converted the age variable to type Int

str(happy) # displayed the structure of the new happy_clean to check if the variable age was successfully mutated
```

Factors were also organized into meaningful order as follows: 
```{r}
#insert relevant R code for Q3 here
```


## Data Exploration 

Two variables of interest are `...` and `...`. Their distributions are as follow: 
```{r}
#insert relevant R code here
#include graphics. DESCRIBE your plot in context. Make sure it includes axis labels and appropriate title. 
```

The distribution for `...` is `....`, suggesting that `....` 




